---
title: Drug-Target Interactions
author: "Authors: Kevin Horan, Yuzhu Duan, Austin Leong, Dan Evans, Jamison McCorrison, Nicholas Schork and Thomas Girke"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  BiocStyle::html_document:
    toc_float: true
    code_folding: show
vignette: |
  %\VignetteIndexEntry{Drug-Target Interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

fontsize: 14pt
bibliography: bibtex.bib
---

<style>
pre code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

<!---
- Compile from command-line
Rscript -e "rmarkdown::render('drugTargetInteractions.Rmd', c('BiocStyle::html_document', 'pdf_document')); knitr::knit('drugTargetInteractions.Rmd', tangle=TRUE)"
-->

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
options(width=100, max.print=1000)
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE")))
```

```{r hidden_setup, echo=FALSE, message=FALSE, warning=FALSE}
library(drugTargetInteractions)
## Small helper: is a UniChem SQLite already built and cached locally?
## Used throughout to skip live examples that would otherwise trigger
## an ~hour-long build on a machine that has never run buildUnichemDb().
.haveUnichemDb <- function() {
    tryCatch({p <- buildUnichemDb(rerun = FALSE); !is.na(p) && file.exists(p)},
             error = function(e) FALSE)
}
```

# Introduction

## Overview

The `drugTargetInteractions` package provides utilities for identifying
**drug-target interactions** starting from *either* a gene/protein identifier
(target &rarr; drug) *or* a compound identifier (drug &rarr; target). It
covers five independent drug-target data sources - [ChEMBL](https://www.ebi.ac.uk/chembl/)
[@Gaulton2012-dv; @Bento2014-ry], [PubChem](https://pubchem.ncbi.nlm.nih.gov/),
[DGIdb](https://dgidb.org/), [Open Targets](https://platform.opentargets.org/)
and the [Therapeutic Target Database (TTD)](https://ttd.idrblab.cn/) - each
queried through its own live REST/GraphQL API or, for TTD, a small local
SQLite built from its public flat files. ChEMBL remains additionally
available through a downloaded local SQLite instance for users who prefer
that (see the [Supplement]).

Because every source has its own native identifier vocabulary (a ChEMBL
molecule ID is not a PubChem CID is not a DrugBank ID), the package also
provides a **generic ID-translation layer**: live services for mapping
protein/gene identifiers ([UniProt](https://www.uniprot.org/)'s REST ID
Mapping API [@Wu2006-jm]) and finding paralogs/orthologs
([Ensembl](https://www.ensembl.org/)'s homology REST API), plus a local
SQLite built from [UniChem](https://www.ebi.ac.uk/unichem/)'s bulk compound
cross-reference table for translating between compound ID types (ChEMBL,
PubChem, DrugBank, ChEBI, and more). A dispatcher function,
`queryDrugTargets()`, ties all of this together: give it an identifier of
*any* recognised type and it resolves that identifier to whatever native ID
each requested source needs, queries one or more sources, and returns the
results.

This vignette is organised as follows:

- **[Quick Start]** - a five-minute tour of the essential commands.
- **[Data Sources in Detail]** - one section per source, with both query
  directions.
- **[ID Translation Layer]** - the standalone translation functions.
- **[Cross-Source Queries]** - `queryDrugTargets()` across one or many
  sources at once, performance guidance, and combining results into a
  single table.
- **[Supplement]** - legacy/alternative functions kept for backward
  compatibility.

## Install Package

As Bioconductor package `drugTargetInteraction` can be installed with the `BiocManager::install()` function.

```{r pkg_install_bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("drugTargetInteractions")
```

Alternatively, the package can be installed from GitHub as follows.

```{r pkg_install_github, eval=FALSE}
devtools::install_github("girke-lab/drugTargetInteractions", build_vignettes=TRUE) # Installs from github
```

## Load Package and Access Help

```{r package_load, eval=TRUE, warning=FALSE}
library("drugTargetInteractions") # Loads the package
```

```{r documentation, eval=FALSE, warning=FALSE}
library(help="drugTargetInteractions") # Lists package info
vignette(topic="drugTargetInteractions", package="drugTargetInteractions") # Opens vignette
```

# Quick Start

This section is meant to be read (and run) in about five minutes. It uses
only sources that need no local database setup, so every chunk here works
immediately after installing the package - no downloads beyond the live API
calls themselves.

## One Source, One Call

Every per-source function uses the same `queryBy = list(molType, idType,
ids)` interface. Target &rarr; drug for the FGFR1 gene via ChEMBL:

```{r quickstart_chembl, eval=TRUE}
getChemblDrugTarget(list(molType = "protein", idType = "Uniprot", ids = "P11362"))[
    , c("QueryIDs", "Drug_Name", "Action_Type", "Max_Phase")]
```

The reverse direction (drug &rarr; target) uses the same function with
`molType = "cmp"`:

```{r quickstart_chembl_reverse, eval=TRUE}
getChemblDrugTarget(list(molType = "cmp", idType = "chembl_id", ids = "CHEMBL1421"))[  # dasatinib
    , c("QueryIDs", "UniProt_ID", "Action_Type")]
```

## Many Sources, One Call

`queryDrugTargets()` accepts an identifier of essentially any type (not
just each source's own native ID - see [ID Translation Layer]) and queries
as many sources as you like in one call:

```{r quickstart_meta, eval=TRUE}
res <- queryDrugTargets(
    list(molType = "gene", idType = "symbol", ids = "FGFR1"),
    sources = c("chembl", "pubchem", "dgidb", "opentargets"))
names(res)              # one data.frame per source that returned results
sapply(res, nrow)
```

## One Table

`res` above is a list - each source keeps its own native columns. To see
one aligned table across sources, use `combineDrugTargets()`:

```{r quickstart_combine, eval=TRUE}
combineDrugTargets(res)[1:6, ]
```

That is the essential workflow: pick an identifier, pick one or more
sources, query, optionally combine. The rest of this vignette covers each
piece in more detail.

# Data Sources in Detail

Each source below is queried through its own bidirectional function, using
that source's *native* identifier types directly. `queryDrugTargets()`
(see [Cross-Source Queries]) builds on top of these by resolving other
identifier types first.

## ChEMBL

ChEMBL [@Gaulton2012-dv; @Bento2014-ry] is queried live through its REST
API via `getChemblDrugTarget()`. Native identifiers: a **UniProt
accession** for target &rarr; drug, a **ChEMBL molecule ID** for drug
&rarr; target.

```{r chembl_target2drug, eval=TRUE}
chembl_t2d <- getChemblDrugTarget(
    list(molType = "protein", idType = "Uniprot", ids = "P11362"))  # FGFR1
head(chembl_t2d[, c("Drug_Name", "MOA", "Action_Type", "Max_Phase")])
```

```{r chembl_drug2target, eval=TRUE}
chembl_d2t <- getChemblDrugTarget(
    list(molType = "cmp", idType = "chembl_id", ids = "CHEMBL25"))  # aspirin
chembl_d2t[, c("UniProt_ID", "Organism", "MOA")]
```

Large ID batches are chunked automatically (see [Performance Considerations]).
Standalone helpers are also available: `getChemblMolecule()` (compound
properties), `getChemblTarget()` (search targets by gene/protein name), and
`getChemblBioactivities()` (raw IC50/Ki/Kd/EC50 measurements for one
target).

## PubChem

PubChem's bioactivity data is queried through its PUG-REST API and NCBI
E-utilities via `getPubchemDrugTarget()`. Native identifiers: a **gene
symbol** (an NCBI GeneID also passes through unresolved) for target &rarr;
drug, a **compound name** (a PubChem CID also passes through unresolved)
for drug &rarr; target.

```{r pubchem_target2drug, eval=TRUE}
pubchem_t2d <- getPubchemDrugTarget(
    list(molType = "gene", idType = "symbol", ids = "KLB"))
head(pubchem_t2d[, c("drug_name", "activity_name", "activity_value_uM")])
```

```{r pubchem_drug2target, eval=TRUE}
pubchem_d2t <- getPubchemDrugTarget(
    list(molType = "cmp", idType = "name", ids = "aspirin"))
pubchem_d2t[, c("gene_symbol", "activity_name", "activity_value_uM")]
```

PubChem is compound-centric and high-volume; results are capped per gene
(`maxCids`, default 400) and filtered to Active, numeric, potency-endpoint
measurements (IC50/Ki/Kd/EC50/AC50/Potency) by default. Because
compound-centric assay data spans whatever species each assay used (e.g.
aspirin's classic COX1/COX2 potency assays are annotated against *sheep*
GeneIDs in PubChem), `getPubchemTargets()`/`getPubchemDrugTarget()` filter
to human (`taxid = 9606`) by default - pass `taxid = NULL` for all species.

## DGIdb

The Drug Gene Interaction Database is queried through its GraphQL API via
`getDgidbDrugTarget()`. DGIdb's schema is bidirectional by construction - a
single `interactions()` query returns nodes that already carry both a gene
and a drug identity - so both directions use the same underlying query,
just filtered by a different argument. Native identifiers: a **gene
symbol** or a **drug name**, both matched case-insensitively.

```{r dgidb_target2drug, eval=TRUE}
dgidb_t2d <- getDgidbDrugTarget(
    list(molType = "gene", idType = "symbol", ids = "FGFR1"))
head(dgidb_t2d[, c("drug_name", "interaction_types", "sources")])
```

```{r dgidb_drug2target, eval=TRUE}
dgidb_d2t <- getDgidbDrugTarget(
    list(molType = "cmp", idType = "name", ids = "imatinib"))
head(dgidb_d2t[, c("gene_name", "interaction_types", "interaction_score")])
```

The `sources` column matters for reuse: DGIdb aggregates interactions from
many upstream databases (ChEMBL, DrugBank, TTD, and others), each with its
own license - a row's `sources` value tells you where that specific
interaction actually came from.

## Open Targets

Open Targets is queried through its GraphQL API via
`getOpenTargetsDrugTarget()`. Unlike DGIdb, its schema is asymmetric:
target &rarr; drug goes through `Target.drugAndClinicalCandidates`, drug
&rarr; target through `Drug.mechanismsOfAction`. Native identifiers: a
**gene symbol** (an Ensembl gene ID also passes through unresolved) for
target &rarr; drug, a **compound name** (a ChEMBL ID also passes through
unresolved, since Open Targets drug IDs *are* ChEMBL IDs) for drug &rarr;
target.

```{r opentargets_target2drug, eval=TRUE}
ot_t2d <- getOpenTargetsDrugTarget(
    list(molType = "gene", idType = "symbol", ids = "FGFR1"))
head(ot_t2d[, c("drug_name", "mechanism_of_action", "max_clinical_stage")])
```

```{r opentargets_drug2target, eval=TRUE}
ot_d2t <- getOpenTargetsDrugTarget(
    list(molType = "cmp", idType = "name", ids = "aspirin"))
ot_d2t[, c("approved_symbol", "mechanism_of_action")]  # PTGS1/PTGS2 = COX1/COX2
```

Note that agonist targets or those without a drugged pocket (e.g. FGF21,
NLRP3, TFEB, ADIPOR1/2 in this package's original test-gene set)
legitimately return zero rows from Open Targets - that reflects the state
of drug discovery for that target, not a query failure.

## TTD (Therapeutic Target Database)

TTD has no live API. Its flat files are downloaded once and assembled into
a small local SQLite via `buildTtdDb()`; `ttdTargetAnnot()` then queries
that local database repeatedly. This mirrors ChEMBL's local-SQLite option
(see the [Supplement]) - build once, query many times.

```{r ttd_build, eval=TRUE}
ttdDbPath <- buildTtdDb(rerun = FALSE)  # builds once, reuses the cached file afterward
```

Native identifiers for `ttdTargetAnnot()`: for target &rarr; drug, a
**gene symbol** (`idType = "symbol"`), a **UniProt mnemonic entry name**
such as `"FGFR1_HUMAN"` (`idType = "uniprot"` - note this is *not* the
UniProt accession `"P11362"` other sources use), or TTD's own
**TargetID** (`idType = "ttd_target_id"`); for drug &rarr; target, a
**drug name** or TTD's own **DrugID** (`idType = "ttd_drug_id"`).

```{r ttd_target2drug, eval=TRUE}
ttd_t2d <- ttdTargetAnnot(
    list(molType = "protein", idType = "symbol", ids = "FGFR1"), ttdDbPath)
head(ttd_t2d[, c("DrugName", "Highest_status", "MOA")])
```

```{r ttd_drug2target, eval=TRUE}
ttd_d2t <- ttdTargetAnnot(
    list(molType = "cmp", idType = "name", ids = "Pemigatinib"), ttdDbPath)
ttd_d2t[, c("GeneName", "Highest_status", "MOA")]
```

**Distribution note:** `buildTtdDb()` only ever downloads TTD's own files
into your local cache and builds the SQLite there - the package itself
ships no TTD data (TTD's license permits academic use but has no explicit
redistribution grant).

# ID Translation Layer

Every source function above expects that source's *native* identifier
type. The functions in this section translate between identifier types
directly, and are also what `queryDrugTargets()` uses internally.

## Protein/Gene IDs

`getUniprotMapping()` maps identifiers via UniProt's REST ID Mapping
service. `from`/`to` use UniProt's own database-name vocabulary (see
[UniProt's ID Mapping help page](https://www.uniprot.org/help/id_mapping)
for the full list) - a few of the most useful:

```{r uniprot_mapping, eval=TRUE}
getUniprotMapping(c("FGFR1", "KLB"), from = "Gene_Name", to = "UniProtKB-Swiss-Prot")
getUniprotMapping("P11362", from = "UniProtKB_AC-ID", to = "Ensembl")
```

`taxId` restricts matches to one organism (default `9606` = human) -
important because a gene symbol can match entries from several species.

## Paralogs and Orthologs

`getEnsemblParalogs()`/`getEnsemblOrthologs()` query Ensembl's homology
REST API. Their `type` classification (`ortholog_one2one`,
`other_paralog`, etc.) comes from gene-tree reconciliation, not raw
sequence-identity thresholding, so it is a more principled "best hit"
signal than picking the highest `perc_id` alone - prefer
`type == "ortholog_one2one"` when you need a single best cross-species
match.

```{r ensembl_paralogs, eval=TRUE}
getEnsemblParalogs("NLRP3")[, c("homolog_id", "type", "perc_id")]
```

```{r ensembl_orthologs, eval=TRUE}
getEnsemblOrthologs("NLRP3", targetSpecies = "mouse")[
    , c("homolog_id", "homolog_species", "type", "perc_id")]
```

Large gene families need the default query shape here - by design,
`getEnsemblParalogs()`/`getEnsemblOrthologs()` already avoid a
pathological slowdown found for such families (see
`?.dot-dtiEnsemblHomologyFetch` for details); pass `condensed = TRUE` for
an even faster existence-check-only path when you don't need
`perc_id`/`perc_pos`.

## Compound IDs

`getUnichemMapping()` translates compound identifiers via a local SQLite
built from UniChem's bulk compound cross-reference table
(`buildUnichemDb()`). It is genuinely source-agnostic: `from`/`to` accept
*any* source name present in the underlying data (not a fixed list
hand-coded per database) - `"chembl"`, `"pubchem"`, `"drugbank"`,
`"chebi"`, and dozens more.

```{r unichem_build_note, eval=FALSE}
## One-time setup (not run automatically in this vignette - takes on the
## order of an hour and downloads ~1.5GB). Rerun = FALSE afterward always
## reuses whatever was built, regardless of which day that was.
unichemDbPath <- buildUnichemDb()
```

```{r unichem_mapping, eval=.haveUnichemDb()}
unichemDbPath <- buildUnichemDb(rerun = FALSE)
getUnichemMapping("CHEMBL25", from = "chembl", to = "pubchem", unichemDbPath)   # aspirin -> CID
getUnichemMapping("CHEMBL25", from = "chembl", to = "drugbank", unichemDbPath) # aspirin -> DrugBank ID
```

Unlike UniProt's or Ensembl's REST APIs, UniChem's own live per-compound
API does not support batch queries and was found to be intermittently
unreliable for even single lookups - the local-SQLite approach exists
specifically to avoid depending on it (see `?buildUnichemDb`). Because the
build is a real, deliberate operation (not something to trigger silently),
`buildUnichemDb(rerun = FALSE)` is always required explicitly before using
`getUnichemMapping()` or `queryDrugTargets()` on structured compound IDs.

# Cross-Source Queries

## Querying One or All Sources

`queryDrugTargets()` resolves `queryBy$ids` to whatever native identifier
each requested `sources` entry needs, then dispatches. `queryBy$idType`
uses the same canonical vocabulary as the translation layer above -
`"symbol"`/`"uniprot"`/`"ensembl"` for genes, `"name"`/`"chembl_id"`/
`"pubchem_id"`/`"drugbank_id"`/`"chebi_id"` for compounds - not each
source's own native type.

```{r meta_one_source, eval=TRUE}
res_one <- queryDrugTargets(
    list(molType = "gene", idType = "symbol", ids = "FGFR1"), sources = "chembl")
names(res_one)
```

```{r meta_ensembl_start, eval=TRUE}
## Starting from an Ensembl gene ID instead of a symbol - resolved
## automatically (via a UniProt accession as an intermediate hop, since
## UniProt's mapping API requires one side of any call to be UniProtKB).
res_ensembl <- queryDrugTargets(
    list(molType = "gene", idType = "ensembl", ids = "ENSG00000077782"),
    sources = c("chembl", "opentargets"))
sapply(res_ensembl, nrow)
```

```{r meta_ttd, eval=TRUE}
## TTD needs a local db path (see above); other sources ignore it.
res_all_live <- queryDrugTargets(
    list(molType = "gene", idType = "symbol", ids = "FGFR1"),
    sources = c("chembl", "pubchem", "dgidb", "opentargets", "ttd"),
    ttdDbPath = ttdDbPath)
sapply(res_all_live, nrow)
```

Starting from a *compound* identifier that needs the UniChem SQLite works
the same way once one has been built:

```{r meta_drugbank_start, eval=.haveUnichemDb()}
res_drugbank <- queryDrugTargets(
    list(molType = "cmp", idType = "drugbank_id", ids = "DB00945"),  # aspirin
    sources = "chembl", unichemDbPath = unichemDbPath)
res_drugbank$chembl[, c("QueryIDs", "UniProt_ID", "Organism")]
```

A source that a query resolved nothing for, or whose own call failed (e.g.
a required local database path was not supplied), simply contributes no
element to the result list rather than raising an error - set
`verbose = TRUE` on any of the calls above to see why.

## Performance Considerations

None of the five sources support unlimited-size ID batches in a single
request, and the live APIs are courtesy-throttled client-side regardless.
The defaults already reflect each source's real constraints (do not raise
these without a specific reason):

| Source | Constraint | Default batch size |
|---|---|---|
| ChEMBL | REST server request-line length (~4KB; `httr2` percent-encodes separators) | 200 IDs/request |
| PubChem | Compound-property lookup batch | 100 CIDs/request; bioactivity summary 50 CIDs/request |
| DGIdb | Courtesy chunking (no documented server limit found) | 300 names/request |
| Open Targets | GraphQL aliasing payload size | 25 targets/request |
| UniProt ID Mapping | Server-enforced job size cap | up to 100,000 IDs/job (asynchronous) |
| UniChem | No batching in the live API at all - hence the local-SQLite design | n/a (local queries) |

All five per-source functions and `queryDrugTargets()` chunk large `ids`
vectors into these batch sizes automatically, so a query of, say, 2,000
UniProt accessions works the same as one - it just takes proportionally
longer and issues proportionally more requests. When testing or debugging
with a deliberately small `chunkSize`, pick one relative to the *endpoint's
fan-out*, not arbitrarily small: a tiny `chunkSize` against a low-fan-out
endpoint (e.g. ChEMBL's `molecule` lookup, one record per ID) is a cheap
way to exercise pagination, but the same tiny `chunkSize` against a
high-fan-out join (e.g. ChEMBL's target/mechanism join, where a handful of
UniProt accessions can expand into 100+ drug rows) multiplies into far
more live requests than intended.

## Combining Results

`combineDrugTargets()` row-binds a small set of canonical columns
(`query_id`, `gene_symbol`, `drug_name`, `action`, `source` by default)
across whichever sources are present in a `queryDrugTargets()` result,
mapping each source's own column names to this shared vocabulary:

```{r combine_default, eval=TRUE}
combined <- combineDrugTargets(res_all_live)
table(combined$source)
head(combined)
```

`query_id` reflects your *original* query token, not each source's own
(possibly translated) `QueryIDs` - useful once a query starts from a
non-native identifier, as in `res_ensembl` above:

```{r combine_query_id, eval=TRUE}
unique(combineDrugTargets(res_ensembl)$query_id)  # still "ENSG00000077782" for every source
```

ChEMBL's REST output has no gene-symbol column at all (it is
UniProt-accession-keyed), so `gene_symbol` is `NA` for ChEMBL rows by
default. Pass `resolveGeneSymbol = TRUE` to fill it in via one extra
`getUniprotMapping()` call:

```{r combine_resolve_symbol, eval=TRUE}
combineDrugTargets(res_all_live, resolveGeneSymbol = TRUE) |>
    subset(source == "ChEMBL") |> head(3)
```

**This is deliberately just column-name alignment and row append** - not
deduplication and not cross-source identity resolution (the same compound
or target appearing under different native IDs in different sources is
not merged into one row). A more principled harmonized container - an S4
class analysed but not yet built, so that different sources' rows for the
literal same compound/target could be properly joined - is a separate,
deferred phase. Every source's full original data remains available
unchanged in the list `queryDrugTargets()` returned, regardless of what
you pass to `combineDrugTargets()`.

# Supplement

The functions below predate, and are independent of, the sources and
translation layer described above. They remain available as
backward-compatible/alternative options but are not the default workflow
recommended by this vignette.

## Local ChEMBL SQLite Workflow

Before ChEMBL's REST API was added, this package's ChEMBL support was
exclusively through a downloaded local SQLite instance of the full ChEMBL
database, queried via `drugTargetAnnot()` (and the related
`getDrugTarget()`, which queries a pre-generated flat-file export rather
than the SQLite directly). This remains useful for bulk, offline, or
very-high-volume ChEMBL querying where the REST API's batch limits (see
[Performance Considerations]) are impractical.

```{r legacy_chembl_download, eval=FALSE}
config <- genConfig(chemblDbPath = "chembldb.db")
downloadChemblDb(rerun = TRUE, config = config)  # downloads the full ChEMBL SQLite (several GB)
queryBy <- list(molType = "protein", idType = "UniProt_ID", ids = "P11362")
drugTargetAnnot(queryBy, config = config)
```

## UniProt.ws-Based ID Mapping

`getUniprotIDs()` predates `getUniprotMapping()` (see [ID Translation
Layer]) and is based on the Bioconductor `r Biocpkg("UniProt.ws")`
package rather than a direct REST call. It is kept for backward
compatibility; `getUniprotMapping()` is the recommended path going
forward.

```{r legacy_uniprot_ws, eval=FALSE}
getUniprotIDs(taxId = 9606, kt = "ENSEMBL", keys = c("ENSG00000077782"))
```

## biomaRt-Based Paralogs

`getParalogs()` predates `getEnsemblParalogs()`/`getEnsemblOrthologs()`
(see [ID Translation Layer]) and is based on `r CRANpkg("biomaRt")`'s
BioMart interface rather than Ensembl's REST API directly. It returns
within-human paralogs only (no cross-species orthologs). Kept for
backward compatibility.

```{r legacy_paralogs, eval=FALSE}
queryBy <- list(molType = "gene", idType = "external_gene_name", ids = "FGFR1")
getParalogs(queryBy)
```

## Static UniChem Mirror

`downloadUniChem()`/`cmpIdMapping()` predate `buildUnichemDb()`/
`getUnichemMapping()` (see [ID Translation Layer]) and work from a small,
static, pre-computed ChEMBL&harr;{DrugBank, PubChem, ChEBI} cross-reference
snapshot hosted on a project S3 bucket, rather than UniChem's own current
bulk data. Kept for backward compatibility and for the local-ChEMBL-SQLite
workflow above, which uses it internally via `cmpIdMapping()`.

```{r legacy_unichem, eval=FALSE}
downloadUniChem(rerun = TRUE)
cmpIdMapping(outfile = "cmp_ids.rds", rerun = TRUE)
```

# Session Info

```{r sessionInfo}
sessionInfo()
```

# References
