---
title: "drugTargetInteractions: Mining the Druggable Genome"
author: "Authors: Kevin Horan, Brandon Le, Brendan Gongol, Desmond Cairo, Mohammad Afsharinia, Tandis Salem, Dan Evans, Nicholas Schork and Thomas Girke"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: yes
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    code_folding: show
vignette: |
  %\VignetteIndexEntry{Drug-Target Interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

fontsize: 14pt
bibliography: bibtex.bib
---

<style>
pre code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

<!---
- Compile from command-line
Rscript -e "rmarkdown::render('drugTargetInteractions.Rmd', c('BiocStyle::html_document', 'pdf_document')); knitr::knit('drugTargetInteractions.Rmd', tangle=TRUE)"
-->

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
options(width=100, max.print=1000)
knitr::opts_chunk$set(
    eval=as.logical(Sys.getenv("KNITR_EVAL", "TRUE")),
    cache=as.logical(Sys.getenv("KNITR_CACHE", "TRUE")))
```

```{r hidden_setup, echo=FALSE, message=FALSE, warning=FALSE}
library(drugTargetInteractions)
## Small helper: is a UniChem SQLite already built and cached locally?
## Used throughout to skip live examples that would otherwise trigger
## an ~hour-long build on a machine that has never run buildUnichemDb().
.haveUnichemDb <- function() {
    tryCatch({p <- buildUnichemDb(rerun = FALSE); !is.na(p) && file.exists(p)},
             error = function(e) FALSE)
}
```

# Introduction

## Overview

The `drugTargetInteractions` package provides utilities for identifying
**drug-target interactions** starting from *either* a gene/protein identifier
(target &rarr; drug) *or* a compound identifier (drug &rarr; target). It
covers seven independent drug-target data sources - [ChEMBL](https://www.ebi.ac.uk/chembl/)
[@Gaulton2012-dv; @Bento2014-ry], [PubChem](https://pubchem.ncbi.nlm.nih.gov/),
[DGIdb](https://dgidb.org/), [Open Targets](https://platform.opentargets.org/),
the [Therapeutic Target Database (TTD)](https://ttd.idrblab.cn/), the
[Broad Institute Drug Repurposing Hub](https://repo-hub.broadinstitute.org/repurposing)
(CLUE) and the [IUPHAR/BPS Guide to PHARMACOLOGY](https://www.guidetopharmacology.org/)
(GtoPdb) - each queried through its own live REST/GraphQL API or, for TTD,
the Repurposing Hub and GtoPdb, a small local SQLite built from public
bulk data. ChEMBL remains additionally available through a downloaded
local SQLite instance for users who prefer that (see the [Supplement]).

Six of these (ChEMBL, DGIdb, Open Targets, TTD, Broad Repurposing Hub,
GtoPdb) provide curated **drug-target annotations** - an expert-curated call
that a drug acts on a target via some mechanism - and are queried the same
way via `queryDrugTargets()`/`combineDrugTargets()` (see [Cross-Source
Queries]). PubChem instead provides raw **bioassay measurements**
(individual assay results, no mechanism claim attached) - a genuinely
different kind of data, covered separately in [Bioassay Queries] alongside
ChEMBL's own bioassay function, `getChemblBioassay()`.

Because every source has its own native identifier vocabulary (a ChEMBL
molecule ID is not a PubChem CID is not a DrugBank ID), the package also
provides a **generic ID-translation layer**: live services for mapping
protein/gene identifiers ([UniProt](https://www.uniprot.org/)'s REST ID
Mapping API [@Wu2006-jm]) and finding paralogs/orthologs
([Ensembl](https://www.ensembl.org/)'s homology REST API), plus a local
SQLite built from [UniChem](https://www.ebi.ac.uk/unichem/)'s bulk compound
cross-reference table for translating between compound ID types (ChEMBL,
PubChem, DrugBank, ChEBI, and more). A dispatcher function,
`queryDrugTargets()`, ties all of this together: give it an identifier of
*any* recognised type and it resolves that identifier to whatever native ID
each requested source needs, queries one or more sources, and returns the
results.

This vignette is organised as follows:

- **[Quick Start]** - a five-minute tour of the essential commands.
- **[Data Sources in Detail]** - one section per annotation source, with
  both query directions.
- **[Bioassay Queries]** - the separate, raw-measurement track (ChEMBL,
  PubChem).
- **[ID Translation Layer]** - the standalone translation functions.
- **[Cross-Source Queries]** - `queryDrugTargets()` across one or many
  sources at once, performance guidance, and combining results into a
  single table.
- **[Genome-Wide Master Table]** - an HGNC-anchored, checkpointed build
  covering all human protein-coding genes at once.
- **[Supplement]** - legacy/alternative functions kept for backward
  compatibility.

## Install Package

As Bioconductor package `drugTargetInteraction` can be installed with the `BiocManager::install()` function.

```{r pkg_install_bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("drugTargetInteractions")
```

Alternatively, the package can be installed from GitHub as follows.

```{r pkg_install_github, eval=FALSE}
if (!requireNamespace("pak", quietly = TRUE))
    install.packages("pak")
pak::pak("girke-lab/drugTargetInteractions") # Installs from github
```

## Load Package and Access Help

```{r package_load, eval=TRUE, warning=FALSE}
library("drugTargetInteractions") # Loads the package
```

```{r documentation, eval=FALSE, warning=FALSE}
library(help="drugTargetInteractions") # Lists package info
vignette(topic="drugTargetInteractions", package="drugTargetInteractions") # Opens vignette
```

# Quick Start

This section is meant to be read (and run) in about five minutes. It uses
only sources that need no local database setup, so every chunk here works
immediately after installing the package - no downloads beyond the live API
calls themselves.

## One Source, One Call

Every per-source function uses the same `queryBy = list(molType, idType,
ids)` interface. Target &rarr; drug for the FGFR1 gene via ChEMBL:

```{r quickstart_chembl, eval=TRUE}
getChemblDrugTarget(list(molType = "protein", idType = "Uniprot", ids = "P11362"))[
    , c("QueryIDs", "Drug_Name", "Action_Type", "Max_Phase")]
```

The reverse direction (drug &rarr; target) uses the same function with
`molType = "cmp"`:

```{r quickstart_chembl_reverse, eval=TRUE}
getChemblDrugTarget(list(molType = "cmp", idType = "chembl_id", ids = "CHEMBL1421"))[  # dasatinib
    , c("QueryIDs", "UniProt_ID", "Action_Type")]
```

## Many Sources, One Call

`queryDrugTargets()` accepts an identifier of essentially any type (not
just each source's own native ID - see [ID Translation Layer]) and queries
as many sources as you like in one call:

```{r quickstart_meta, eval=TRUE}
res <- queryDrugTargets(
    list(molType = "gene", idType = "symbol", ids = "FGFR1"),
    sources = c("chembl", "dgidb", "opentargets"))
names(res)              # one data.frame per source that returned results
sapply(res, nrow)
```

## One Table

`res` above is a list - each source keeps its own native columns. To see
one aligned table across sources, use `combineDrugTargets()`:

```{r quickstart_combine, eval=TRUE}
combineDrugTargets(res)[1:6, ]
```

That is the essential workflow: pick an identifier, pick one or more
sources, query, optionally combine. The rest of this vignette covers each
piece in more detail.

# Data Sources in Detail

Each source below is queried through its own bidirectional function, using
that source's *native* identifier types directly. `queryDrugTargets()`
(see [Cross-Source Queries]) builds on top of these by resolving other
identifier types first.

The table below is a quick-reference cheat sheet for every valid
`queryBy = list(molType, idType, ids)` combination accepted by these
per-source functions - each row corresponds to one working call, with a
real, live-tested example value for `ids`. Full detail on each source
follows in the subsections below.

| Function                       | `molType`           | `idType`         | Example `ids`               |
|---------------------------------|----------------------|-------------------|------------------------------|
| `getChemblDrugTarget()`         | `"protein"`          | `"Uniprot"`       | `"P11362"` (FGFR1)          |
| `getChemblDrugTarget()`         | `"cmp"`              | `"chembl_id"`     | `"CHEMBL25"` (aspirin)      |
| `getDgidbDrugTarget()`          | `"gene"`             | `"symbol"`        | `"FGFR1"`                   |
| `getDgidbDrugTarget()`          | `"cmp"`              | `"name"`          | `"imatinib"`                |
| `getOpenTargetsDrugTarget()`    | `"gene"`             | `"symbol"`        | `"FGFR1"`                   |
| `getOpenTargetsDrugTarget()`    | `"cmp"`              | `"name"`          | `"aspirin"`                 |
| `getPubchemDrugTarget()`        | `"gene"`             | `"symbol"`        | `"KLB"`                     |
| `getPubchemDrugTarget()`        | `"cmp"`              | `"name"`          | `"aspirin"`                 |
| `ttdTargetAnnot()`              | `"protein"`          | `"symbol"`        | `"FGFR1"`                   |
| `ttdTargetAnnot()`              | `"protein"`          | `"uniprot"`       | `"FGFR1_HUMAN"` (mnemonic, not accession) |
| `ttdTargetAnnot()`              | `"protein"`          | `"ttd_target_id"` | `"T47101"`                  |
| `ttdTargetAnnot()`              | `"cmp"`              | `"name"`          | `"Pemigatinib"`             |
| `ttdTargetAnnot()`              | `"cmp"`              | `"ttd_drug_id"`   | `"D0O6UY"`                  |
| `broadRepurposingHubAnnot()`    | `"protein"`/`"gene"` | `"symbol"`        | `"FGFR1"`                   |
| `broadRepurposingHubAnnot()`    | `"cmp"`              | `"name"`          | `"erdafitinib"`             |
| `broadRepurposingHubAnnot()`    | `"cmp"`              | `"broad_id"`      | `"BRD-K84868168-001-01-0"`  |
| `gtoPdbTargetAnnot()`           | `"protein"`/`"gene"` | `"symbol"`        | `"FGFR1"`                   |
| `gtoPdbTargetAnnot()`           | `"protein"`/`"gene"` | `"gtp_target_id"` | `"1808"`                    |
| `gtoPdbTargetAnnot()`           | `"cmp"`              | `"name"`          | `"pemigatinib"`             |
| `gtoPdbTargetAnnot()`           | `"cmp"`              | `"gtp_ligand_id"` | `"9767"` (pemigatinib)      |

`ids` accepts a vector of more than one value for every row above; a
single example is shown here for brevity. `getChemblBioassay()` and
`getPubchemDrugTarget()`'s bioassay-track sibling functions (see
[Bioassay Queries]) accept the same `molType`/`idType` combinations as
their annotation counterparts in this table.

## ChEMBL

ChEMBL [@Gaulton2012-dv; @Bento2014-ry] is queried live through its REST
API via `getChemblDrugTarget()`. Native identifiers: a **UniProt
accession** for target &rarr; drug, a **ChEMBL molecule ID** for drug
&rarr; target.

```{r chembl_target2drug, eval=TRUE}
chembl_t2d <- getChemblDrugTarget(
    list(molType = "protein", idType = "Uniprot", ids = "P11362"))  # FGFR1
head(chembl_t2d[, c("Drug_Name", "MOA", "Action_Type", "Max_Phase")])
```

```{r chembl_drug2target, eval=TRUE}
chembl_d2t <- getChemblDrugTarget(
    list(molType = "cmp", idType = "chembl_id", ids = "CHEMBL25"))  # aspirin
chembl_d2t[, c("UniProt_ID", "Organism", "MOA")]
```

Large ID batches are chunked automatically (see [Performance Considerations]).
Standalone helpers are also available: `getChemblMolecule()` (compound
properties), `getChemblTarget()` (search targets by gene/protein name), and
`getChemblBioactivities()` (raw IC50/Ki/Kd/EC50 measurements for one
target).

## DGIdb

The Drug Gene Interaction Database is queried through its GraphQL API via
`getDgidbDrugTarget()`. DGIdb's schema is bidirectional by construction - a
single `interactions()` query returns nodes that already carry both a gene
and a drug identity - so both directions use the same underlying query,
just filtered by a different argument. Native identifiers: a **gene
symbol** or a **drug name**, both matched case-insensitively.

```{r dgidb_target2drug, eval=TRUE}
dgidb_t2d <- getDgidbDrugTarget(
    list(molType = "gene", idType = "symbol", ids = "FGFR1"))
head(dgidb_t2d[, c("drug_name", "interaction_types", "sources")])
```

```{r dgidb_drug2target, eval=TRUE}
dgidb_d2t <- getDgidbDrugTarget(
    list(molType = "cmp", idType = "name", ids = "imatinib"))
head(dgidb_d2t[, c("gene_name", "interaction_types", "interaction_score")])
```

The `sources` column matters for reuse: DGIdb aggregates interactions from
many upstream databases (ChEMBL, DrugBank, TTD, and others), each with its
own license - a row's `sources` value tells you where that specific
interaction actually came from.

## Open Targets

Open Targets is queried through its GraphQL API via
`getOpenTargetsDrugTarget()`. Unlike DGIdb, its schema is asymmetric:
target &rarr; drug goes through `Target.drugAndClinicalCandidates`, drug
&rarr; target through `Drug.mechanismsOfAction`. Native identifiers: a
**gene symbol** (an Ensembl gene ID also passes through unresolved) for
target &rarr; drug, a **compound name** (a ChEMBL ID also passes through
unresolved, since Open Targets drug IDs *are* ChEMBL IDs) for drug &rarr;
target.

```{r opentargets_target2drug, eval=TRUE}
ot_t2d <- getOpenTargetsDrugTarget(
    list(molType = "gene", idType = "symbol", ids = "FGFR1"))
head(ot_t2d[, c("drug_name", "mechanism_of_action", "max_clinical_stage")])
```

```{r opentargets_drug2target, eval=TRUE}
ot_d2t <- getOpenTargetsDrugTarget(
    list(molType = "cmp", idType = "name", ids = "aspirin"))
ot_d2t[, c("approved_symbol", "mechanism_of_action")]  # PTGS1/PTGS2 = COX1/COX2
```

Note that agonist targets or those without a drugged pocket (e.g. FGF21,
NLRP3, TFEB, ADIPOR1/2 in this package's original test-gene set)
legitimately return zero rows from Open Targets - that reflects the state
of drug discovery for that target, not a query failure.

## Requesting Additional Fields

`getChemblDrugTarget()`, `getPubchemDrugTarget()`, `getDgidbDrugTarget()`
and `getOpenTargetsDrugTarget()` return a small, harmonized set of columns
by default (`fields = "core"`) - this is exactly what lets
`combineDrugTargets()` bind rows from different sources into one table
(see [Combining Results]). For a single-source query, though, each
underlying API exposes many more fields than this curated set. Pass
`fields = "all"` to get everything available, source-prefixed to avoid
name clashes between sources:

```{r fields_all, eval=TRUE}
chembl_all <- getChemblDrugTarget(
    list(molType = "cmp", idType = "chembl_id", ids = "CHEMBL25"), fields = "all")
dim(chembl_all)
chembl_all[, c("Drug_Name", "mechanism.mechanism_comment", "molecule.molecule_type")]
```

Use `listDrugTargetFields()` to browse what a source offers before
deciding what to keep:

```{r fields_list, eval=TRUE}
listDrugTargetFields("chembl")[1:10]
```

`fields` also accepts a character vector directly, so a single call can
ask for just the extra columns you actually want:

```{r fields_custom, eval=TRUE}
getChemblDrugTarget(
    list(molType = "cmp", idType = "chembl_id", ids = "CHEMBL25"),
    fields = c("Drug_Name", "molecule.molecule_type", "target.organism"))
```

The same `fields` argument works identically on all four source
functions (and on `queryDrugTargets()`, which forwards it through). For
ChEMBL and PubChem (REST APIs), `fields = "all"` costs nothing extra -
the full response is already fetched, just no longer discarded. For
DGIdb and Open Targets (GraphQL APIs), it does request a wider response,
since GraphQL only returns fields a query explicitly asks for.

## TTD (Therapeutic Target Database)

TTD has no live API. Its flat files are downloaded once and assembled into
a small local SQLite via `buildTtdDb()`; `ttdTargetAnnot()` then queries
that local database repeatedly. This mirrors ChEMBL's local-SQLite option
(see the [Supplement]) - build once, query many times.

```{r ttd_build, eval=TRUE}
ttdDbPath <- buildTtdDb(rerun = TRUE)  # first build must fetch; rerun = FALSE reuses the cache afterward
```

Native identifiers for `ttdTargetAnnot()`: for target &rarr; drug, a
**gene symbol** (`idType = "symbol"`), a **UniProt mnemonic entry name**
such as `"FGFR1_HUMAN"` (`idType = "uniprot"` - note this is *not* the
UniProt accession `"P11362"` other sources use), or TTD's own
**TargetID** (`idType = "ttd_target_id"`); for drug &rarr; target, a
**drug name** or TTD's own **DrugID** (`idType = "ttd_drug_id"`).

The result's `Uniprot_acc` column carries a resolved UniProt accession
where available (via the package's UniProt REST scaffold, keyed on
`GeneName`, not on string-mangling the mnemonic) - check the paired
`uniprot_source` column before trusting it: `"resolved"` means it
worked, `"TTD_missing"` means TTD itself had no usable ID for that
target at all (overwhelmingly family/pathway-level "targets" with no
single accession by construction, e.g. "Fibroblast growth factor
receptor (FGFR)" rather than the single-protein "FGFR1"), and
`"unresolved"`/`"resolution_failed"` mean resolution was attempted but
failed - a real gene symbol with no UniProt mapping, or a transient
UniProt outage, respectively.

```{r ttd_target2drug, eval=TRUE}
ttd_t2d <- ttdTargetAnnot(
    list(molType = "protein", idType = "symbol", ids = "FGFR1"), ttdDbPath)
head(ttd_t2d[, c("DrugName", "Highest_status", "MOA", "Uniprot_acc", "molecule_type")])
```

```{r ttd_drug2target, eval=TRUE}
ttd_d2t <- ttdTargetAnnot(
    list(molType = "cmp", idType = "name", ids = "Pemigatinib"), ttdDbPath)
ttd_d2t[, c("GeneName", "Highest_status", "MOA", "PubChem_CID", "Indication")]
```

Note `Indication` packs *every* disease Pemigatinib is associated with,
each with its own clinical status (e.g. approved for cholangiocarcinoma
but only Phase 2 for bladder cancer) - richer than the single
`Highest_status` column, which only ever reflects the drug's single
most-advanced status across all diseases combined.

`molecule_type` is a derived classification - TTD's own `DRUGTYPE` field
where available, a SMILES-presence/name heuristic otherwise (see
`?buildTtdDb`) - not an authoritative structural determination. Useful
for filtering before a downstream cheminformatics step (e.g. keep only
`"small molecule"` rows before computing descriptors), but not something
to trust for anything decision-critical without checking `Smiles`
directly.

Like the four REST-backed sources, `ttdTargetAnnot()` supports
`fields = "core"/"all"/<vector>` (see [Requesting Additional Fields]) -
though since `ttd_interactions` is one flat local table with nothing
larger to opt into, `"core"` and `"all"` return the same full column set
here; the option mainly exists to narrow down to a specific subset:

```{r ttd_fields, eval=TRUE}
listDrugTargetFields("ttd")
ttdTargetAnnot(
    list(molType = "protein", idType = "symbol", ids = "FGFR1"), ttdDbPath,
    fields = c("DrugName", "CAS", "ChEBI_ID"))[1:3, ]
```

Every downloaded TTD file's own release stamp (version and date - TTD
does not always move these together across files, see `?buildTtdDb`) is
attached to `ttdTargetAnnot()`'s result as `attr(result, "ttd_release")`:

```{r ttd_release, eval=TRUE}
attr(ttd_t2d, "ttd_release")
```

**Distribution note:** `buildTtdDb()` only ever downloads TTD's own files
into your local cache and builds the SQLite there - the package itself
ships no TTD data (TTD's license permits academic use but has no explicit
redistribution grant). The cross-matching columns (`PubChem_CID`,
`PubChem_SID`, `CAS`, `ChEBI_ID`) are TTD's own; this release's
cross-matching file does not carry `ChEMBL_ID`, `DrugBank_ID`, or
`InChIKey` at all (confirmed by reading the raw file, not just its
header) - joining to those needs a separate hop through
`getUnichemMapping()`/`buildUnichemDb()` via `PubChem_CID`, not this
function.

## Broad Repurposing Hub (CLUE)

Like TTD, the [Broad Institute Drug Repurposing
Hub](https://repo-hub.broadinstitute.org/repurposing) has no live API - two
flat TSV files (drug-level and physical-sample-level annotation) are
downloaded once and assembled into a small local SQLite via
`buildBroadRepurposingHubDb()`; `broadRepurposingHubAnnot()` then queries
that local database repeatedly.

```{r broad_build, eval=TRUE, warning=FALSE, message=FALSE}
brhDbPath <- buildBroadRepurposingHubDb(rerun = TRUE)  # first build must fetch; rerun = FALSE reuses the cache afterward
```

Native identifiers for `broadRepurposingHubAnnot()`: for target &rarr; drug,
a **gene symbol** (`idType = "symbol"` - the Repurposing Hub only exposes
gene symbols, no accession system of its own); for drug &rarr; target, a
**drug name** (`idType = "name"`, matched case-insensitively against the
Hub's own lower-case `pert_iname`) or a specific physical sample/batch ID
(`idType = "broad_id"`).

```{r broad_target2drug, eval=TRUE}
broad_t2d <- broadRepurposingHubAnnot(
    list(molType = "protein", idType = "symbol", ids = "FGFR1"), brhDbPath)
head(broad_t2d[, c("pert_iname", "clinical_phase", "moa")])
```

```{r broad_drug2target, eval=TRUE}
broad_d2t <- broadRepurposingHubAnnot(
    list(molType = "cmp", idType = "name", ids = "pemigatinib"), brhDbPath)
broad_d2t[, c("target_gene", "clinical_phase", "moa")]
```

**Distribution note:** `buildBroadRepurposingHubDb()` only ever downloads
the Repurposing Hub's own files into your local cache and builds the SQLite
there - the package itself ships no Repurposing Hub data. Its license is
more restrictive than TTD's: the source files explicitly state they are
"provided for non-commercial use only."

**Two tables, not one:** unlike TTD, the Repurposing Hub's sample-level
file has a genuinely different grain than a drug-target edge - one
compound routinely has several physical samples/lots. `broad_interactions`
(what `broadRepurposingHubAnnot()` returns above) is kept strictly at
one row per `(pert_iname, target_gene)`; physical-sample/QC metadata
(purity, vendor, catalog number, etc.) lives separately in
`broad_samples`, queryable directly by `broad_id`. A `structure_ambiguous`
column flags the rare case (~2% of compounds) where one `pert_iname`
display name actually covers more than one distinct chemical structure
(different salts/stereoisomers) - `smiles`/`InChIKey`/`pubchem_cid` are
then `"; "`-joined rather than picking one arbitrarily; structure-sensitive
work should key on `InChIKey`, not name, for those rows.

**Known operational caveat, worked around automatically:**
`repo-hub.broadinstitute.org` has been observed serving an incomplete TLS
certificate chain (missing the InCommon intermediate certificate).
Browsers usually mask this by auto-fetching the missing intermediate;
`curl`/R's default download methods generally do not. This is a
server-side issue, not something fixable at the source - but
`downloadBroadRepurposingHub()` now retries automatically with a CA
bundle extended to include the one missing (legitimate, publicly-issued)
certificate, so a plain call should succeed regardless.

## GtoPdb (IUPHAR/BPS Guide to PHARMACOLOGY)

GtoPdb offers a documented REST API, but its per-target endpoints are
one-resource-per-call with no bulk target-to-gene-symbol mapping endpoint -
impractical for a genome-wide build. Instead, `buildGtoPdbDb()` uses
GtoPdb's *bulk* `/services/interactions` endpoint (one call returns the
whole ~24,000-row interaction table as JSON) joined locally against a
small bulk flat file GtoPdb separately publishes for target-ID-to-HGNC
mapping, assembled into a small local SQLite - the same build-once,
query-many pattern as TTD and the Repurposing Hub, just sourced from
REST plus one small file instead of pure flat files.

```{r gtopdb_build, eval=TRUE}
gtoPdbDbPath <- buildGtoPdbDb(rerun = TRUE)  # first build must fetch; rerun = FALSE reuses the cache afterward
```

Native identifiers for `gtoPdbTargetAnnot()`: for target &rarr; drug, a
**gene symbol** (`idType = "symbol"`) or GtoPdb's own **target ID**
(`idType = "gtp_target_id"`); for drug &rarr; target, a **ligand name**
(`idType = "name"`) or GtoPdb's own **ligand ID** (`idType = "gtp_ligand_id"`).
Interactions are pre-filtered to `species == "Human"` at build time.

```{r gtopdb_target2drug, eval=TRUE}
gtopdb_t2d <- gtoPdbTargetAnnot(
    list(molType = "protein", idType = "symbol", ids = "FGFR1"), gtoPdbDbPath)
head(gtopdb_t2d[, c("ligandName", "type", "action", "affinity")])
```

```{r gtopdb_drug2target, eval=TRUE}
gtopdb_d2t <- gtoPdbTargetAnnot(
    list(molType = "cmp", idType = "name", ids = "pemigatinib"), gtoPdbDbPath)
gtopdb_d2t[, c("target_gene", "type", "action", "affinity")]
```

GtoPdb is a more targeted, expert-curated resource than TTD/the
Repurposing Hub, not exhaustive - of this vignette's 8 target genes, only
FGFR1 and NLRP3 are covered.

**Distribution note:** unlike TTD and the Repurposing Hub, GtoPdb's data
is licensed under the Open Data Commons Open Database License (ODbL)
with contents under CC-BY-SA 4.0 - clear terms that explicitly permit
redistribution (with attribution/share-alike). `buildGtoPdbDb()` still
only ever downloads into your local cache and builds the SQLite there,
matching the other local-SQLite sources' posture for consistency, though
that default isn't a licensing requirement here the way it is for TTD/
the Repurposing Hub.

# Bioassay Queries

Everything in [Data Sources in Detail] returns curated drug-target
*annotations* - a database's own expert-curated call that a drug acts on
a target via some mechanism (ChEMBL's `drug_mechanism`, DGIdb's
aggregated interactions, Open Targets' `mechanismsOfAction`, TTD's
target-drug mappings). That's a genuinely different kind of data from a
raw *bioassay measurement* - one experimental result, e.g. "this compound
inhibited this target with an IC50 of 42nM in this specific assay" - with
no claim about mechanism or curation attached. PubChem's REST API only
ever returns the latter, and ChEMBL exposes both kinds side by side. To
keep the two from being silently conflated, this package always keeps
them in separate functions and never combines them:
`queryDrugTargets()`/`combineDrugTargets()` (see [Cross-Source Queries])
only ever touch the four annotation sources - PubChem is not one of
their `sources`.

## ChEMBL

`getChemblBioassay()` is the bioassay-track sibling of
`getChemblDrugTarget()`: same bidirectional `queryBy` interface (UniProt
accession for target &rarr; drug, ChEMBL molecule ID for drug &rarr;
target), but pulling from ChEMBL's `activity` resource instead of
`drug_mechanism`.

```{r chembl_bioassay_target2drug, eval=TRUE}
chembl_ba_t2d <- getChemblBioassay(
    list(molType = "protein", idType = "Uniprot", ids = "P11362"),  # FGFR1
    standardType = "IC50")
head(chembl_ba_t2d[, c("Drug_Name", "standard_value", "standard_units")])
```

```{r chembl_bioassay_drug2target, eval=TRUE}
chembl_ba_d2t <- getChemblBioassay(
    list(molType = "cmp", idType = "chembl_id", ids = "CHEMBL1421"))  # dasatinib
head(chembl_ba_d2t[, c("UniProt_ID", "standard_type", "standard_value")])
```

A single-target convenience helper, `getChemblBioactivities()`, is also
available for a quick one-target lookup without the batching/direction
machinery. `getChemblBioassay()` also supports the same `fields =
"core"/"all"/<vector>` mechanism as the annotation functions (see
[Requesting Additional Fields]) - use `listBioassayFields("chembl")` to
browse the extra columns available:

```{r chembl_bioassay_fields, eval=TRUE}
listBioassayFields("chembl")[1:10]
```

## PubChem

PubChem's bioactivity data is queried through its PUG-REST API and NCBI
E-utilities via `getPubchemDrugTarget()`. Native identifiers: a **gene
symbol** (an NCBI GeneID also passes through unresolved) for target &rarr;
drug, a **compound name** (a PubChem CID also passes through unresolved)
for drug &rarr; target.

```{r pubchem_target2drug, eval=TRUE}
pubchem_t2d <- getPubchemDrugTarget(
    list(molType = "gene", idType = "symbol", ids = "KLB"))
head(pubchem_t2d[, c("drug_name", "activity_name", "activity_value_uM")])
```

```{r pubchem_drug2target, eval=TRUE}
pubchem_d2t <- getPubchemDrugTarget(
    list(molType = "cmp", idType = "name", ids = "aspirin"))
pubchem_d2t[, c("gene_symbol", "activity_name", "activity_value_uM")]
```

PubChem is compound-centric and high-volume; results are capped per gene
(`maxCids`, default 400) and filtered to Active, numeric, potency-endpoint
measurements (IC50/Ki/Kd/EC50/AC50/Potency) by default. Because
compound-centric assay data spans whatever species each assay used (e.g.
aspirin's classic COX1/COX2 potency assays are annotated against *sheep*
GeneIDs in PubChem), `getPubchemTargets()`/`getPubchemDrugTarget()` filter
to human (`taxid = 9606`) by default - pass `taxid = NULL` for all species.
Like ChEMBL's bioassay function, `fields = "core"/"all"/<vector>` and
`listBioassayFields("pubchem")` are available here too.

# ID Translation Layer

Every source function above expects that source's *native* identifier
type. The functions in this section translate between identifier types
directly, and are also what `queryDrugTargets()` uses internally.

## Protein/Gene IDs

`getUniprotMapping()` maps identifiers via UniProt's REST ID Mapping
service. `from`/`to` use UniProt's own database-name vocabulary (see
[UniProt's ID Mapping help page](https://www.uniprot.org/help/id_mapping)
for the full list) - a few of the most useful:

```{r uniprot_mapping, eval=TRUE}
getUniprotMapping(c("FGFR1", "KLB"), from = "Gene_Name", to = "UniProtKB-Swiss-Prot")
getUniprotMapping("P11362", from = "UniProtKB_AC-ID", to = "Ensembl")
```

`taxId` restricts matches to one organism (default `9606` = human) -
important because a gene symbol can match entries from several species.

## Paralogs and Orthologs

`getEnsemblParalogs()`/`getEnsemblOrthologs()` query Ensembl's homology
REST API. Their `type` classification (`ortholog_one2one`,
`other_paralog`, etc.) comes from gene-tree reconciliation, not raw
sequence-identity thresholding, so it is a more principled "best hit"
signal than picking the highest `perc_id` alone - prefer
`type == "ortholog_one2one"` when you need a single best cross-species
match.

```{r ensembl_paralogs, eval=TRUE}
getEnsemblParalogs("NLRP3")[, c("homolog_id", "type", "perc_id")]
```

```{r ensembl_orthologs, eval=TRUE}
getEnsemblOrthologs("NLRP3", targetSpecies = "mouse")[
    , c("homolog_id", "homolog_species", "type", "perc_id")]
```

Large gene families need the default query shape here - by design,
`getEnsemblParalogs()`/`getEnsemblOrthologs()` already avoid a
pathological slowdown found for such families (see
`?.dot-dtiEnsemblHomologyFetch` for details); pass `condensed = TRUE` for
an even faster existence-check-only path when you don't need
`perc_id`/`perc_pos`.

## Compound IDs

`getUnichemMapping()` translates compound identifiers via a local SQLite
built from UniChem's bulk compound cross-reference table
(`buildUnichemDb()`). It is genuinely source-agnostic: `from`/`to` accept
*any* source name present in the underlying data (not a fixed list
hand-coded per database) - `"chembl"`, `"pubchem"`, `"drugbank"`,
`"chebi"`, and dozens more.

```{r unichem_build_note, eval=FALSE}
## One-time setup (not run automatically in this vignette - takes on the
## order of an hour and downloads ~1.5GB). Rerun = FALSE afterward always
## reuses whatever was built, regardless of which day that was.
unichemDbPath <- buildUnichemDb()
```

```{r unichem_mapping, eval=.haveUnichemDb()}
unichemDbPath <- buildUnichemDb(rerun = FALSE)
getUnichemMapping("CHEMBL25", from = "chembl", to = "pubchem", unichemDbPath)   # aspirin -> CID
getUnichemMapping("CHEMBL25", from = "chembl", to = "drugbank", unichemDbPath) # aspirin -> DrugBank ID
```

Unlike UniProt's or Ensembl's REST APIs, UniChem's own live per-compound
API does not support batch queries and was found to be intermittently
unreliable for even single lookups - the local-SQLite approach exists
specifically to avoid depending on it (see `?buildUnichemDb`). Because the
build is a real, deliberate operation (not something to trigger silently),
`buildUnichemDb(rerun = FALSE)` is always required explicitly before using
`getUnichemMapping()` or `queryDrugTargets()` on structured compound IDs.

# Cross-Source Queries

## Querying One or All Sources

`queryDrugTargets()` resolves `queryBy$ids` to whatever native identifier
each requested `sources` entry needs, then dispatches. `queryBy$idType`
uses the same canonical vocabulary as the translation layer above - not
each source's own native type (see the table in [Data Sources in Detail]
for those):

| `molType`            | `idType`        | Example `ids`             |
|-----------------------|-------------------|-----------------------------|
| `"gene"`/`"protein"`  | `"symbol"`        | `"FGFR1"`                  |
| `"gene"`/`"protein"`  | `"uniprot"`       | `"P11362"` (FGFR1)         |
| `"gene"`/`"protein"`  | `"ensembl"`       | `"ENSG00000077782"` (FGFR1)|
| `"cmp"`               | `"name"`          | `"aspirin"`                |
| `"cmp"`               | `"chembl_id"`     | `"CHEMBL25"` (aspirin)     |
| `"cmp"`               | `"pubchem_id"`    | PubChem CID                |
| `"cmp"`               | `"drugbank_id"`   | `"DB00945"` (aspirin)      |
| `"cmp"`               | `"chebi_id"`      | ChEBI ID                   |

Compound-side resolution needs a local UniChem SQLite (`unichemDbPath`,
see [Compound IDs]) for most `idType != to` combinations, except three
zero-network-database shortcuts: `idType`/`to` already identical (a
passthrough, including `"name"` -> `"name"`), `"chembl_id"` -> `"name"`
(direct via `getChemblMolecule()`), and `"name"` -> `"pubchem_id"`
(direct via PubChem's own name search). Every other combination needs
it, including `"name"` -> any structured type *other than*
`"pubchem_id"` (an extra hop past the PubChem CID) and any structured
type *other than* `"chembl_id"` -> `"name"` (an extra hop through a
ChEMBL ID first). Gene/protein-side resolution never needs a local
database - it always goes through UniProt's REST API.

```{r meta_one_source, eval=TRUE}
res_one <- queryDrugTargets(
    list(molType = "gene", idType = "symbol", ids = "FGFR1"), sources = "chembl")
names(res_one)
```

```{r meta_ensembl_start, eval=TRUE}
## Starting from an Ensembl gene ID instead of a symbol - resolved
## automatically (via a UniProt accession as an intermediate hop, since
## UniProt's mapping API requires one side of any call to be UniProtKB).
res_ensembl <- queryDrugTargets(
    list(molType = "gene", idType = "ensembl", ids = "ENSG00000077782"),
    sources = c("chembl", "opentargets"))
sapply(res_ensembl, nrow)
```

```{r meta_ttd, eval=TRUE}
## TTD, Broad Repurposing Hub and GtoPdb each need a local db path (see
## above); other sources ignore the ones they don't use.
res_all_live <- queryDrugTargets(
    list(molType = "gene", idType = "symbol", ids = "FGFR1"),
    sources = c("chembl", "dgidb", "opentargets", "ttd", "broad", "gtopdb"),
    ttdDbPath = ttdDbPath, brhDbPath = brhDbPath, gtoPdbDbPath = gtoPdbDbPath)
sapply(res_all_live, nrow)
```

Starting from a *compound* identifier that needs the UniChem SQLite works
the same way once one has been built:

```{r meta_drugbank_start, eval=.haveUnichemDb()}
res_drugbank <- queryDrugTargets(
    list(molType = "cmp", idType = "drugbank_id", ids = "DB00945"),  # aspirin
    sources = "chembl", unichemDbPath = unichemDbPath)
res_drugbank$chembl[, c("QueryIDs", "UniProt_ID", "Organism")]
```

A source that a query resolved nothing for, or whose own call failed (e.g.
a required local database path was not supplied), simply contributes no
element to the result list rather than raising an error - set
`verbose = TRUE` on any of the calls above to see why.

## Performance Considerations

None of the seven sources support unlimited-size ID batches in a single
request, and the live APIs are courtesy-throttled client-side regardless.
The defaults already reflect each source's real constraints (do not raise
these without a specific reason):

| Source | Constraint | Default batch size |
|---|---|---|
| ChEMBL | REST server request-line length (~4KB; `httr2` percent-encodes separators) | 200 IDs/request |
| PubChem | Compound-property lookup batch | 100 CIDs/request; bioactivity summary 50 CIDs/request |
| DGIdb | Courtesy chunking (no documented server limit found) | 300 names/request |
| Open Targets | GraphQL aliasing payload size | 25 targets/request |
| UniProt ID Mapping | Server-enforced job size cap | up to 100,000 IDs/job (asynchronous) |
| UniChem | No batching in the live API at all - hence the local-SQLite design | n/a (local queries) |

Every per-source function chunks large `ids` vectors into these batch
sizes automatically - including `queryDrugTargets()`, for the six
annotation sources it dispatches to (see [Bioassay Queries] for why
PubChem isn't one of them) - so a query of, say, 2,000 UniProt
accessions works the same as one - it just takes proportionally longer
and issues proportionally more requests. When testing or debugging
with a deliberately small `chunkSize`, pick one relative to the *endpoint's
fan-out*, not arbitrarily small: a tiny `chunkSize` against a low-fan-out
endpoint (e.g. ChEMBL's `molecule` lookup, one record per ID) is a cheap
way to exercise pagination, but the same tiny `chunkSize` against a
high-fan-out join (e.g. ChEMBL's target/mechanism join, where a handful of
UniProt accessions can expand into 100+ drug rows) multiplies into far
more live requests than intended.

## Combining Results

`combineDrugTargets()` row-binds a small set of canonical columns
(`query_id`, `gene_symbol`, `drug_name`, `action`, `source` by default)
across whichever sources are present in a `queryDrugTargets()` result,
mapping each source's own column names to this shared vocabulary:

```{r combine_default, eval=TRUE}
combined <- combineDrugTargets(res_all_live)
table(combined$source)
head(combined)
```

`query_id` reflects your *original* query token, not each source's own
(possibly translated) `QueryIDs` - useful once a query starts from a
non-native identifier, as in `res_ensembl` above:

```{r combine_query_id, eval=TRUE}
unique(combineDrugTargets(res_ensembl)$query_id)  # still "ENSG00000077782" for every source
```

ChEMBL's REST output has no gene-symbol column at all (it is
UniProt-accession-keyed), so `gene_symbol` is `NA` for ChEMBL rows by
default. Pass `resolveGeneSymbol = TRUE` to fill it in via one extra
`getUniprotMapping()` call:

```{r combine_resolve_symbol, eval=TRUE}
combineDrugTargets(res_all_live, resolveGeneSymbol = TRUE) |>
    subset(source == "ChEMBL") |> head(3)
```

**This is deliberately just column-name alignment and row append** - not
deduplication and not cross-source identity resolution (the same compound
or target appearing under different native IDs in different sources is
not merged into one row). A more principled harmonized container - an S4
class analysed but not yet built, so that different sources' rows for the
literal same compound/target could be properly joined - is a separate,
deferred phase. Every source's full original data remains available
unchanged in the list `queryDrugTargets()` returned, regardless of what
you pass to `combineDrugTargets()`.

# Genome-Wide Master Table

`queryDrugTargets()` is designed for on-demand queries of a handful to a
few thousand identifiers. A different, common need is a **periodically
rebuilt, shareable master table** covering essentially all human
protein-coding genes at once - the natural resource for genome-wide
questions like "do disease-associated gene variants have known drugs
annotated?" or "do hits from a LINCS perturbation-signature search have
known targets/MOAs?". A single **target-centric** genome-wide run
already answers the reverse **drug &rarr; target** question too: every
row already carries whichever drug matched that gene, so filtering the
assembled table by drug identifier gives you that direction for free -
no separate compound-centric run is needed.

The code in this section is illustrative only (`eval=FALSE`) - a
genome-wide run queries live APIs for ~19,200 genes and can take hours;
it has no place executing during vignette or `R CMD check` builds. Run
it directly in an interactive session or a scheduled job instead.

## The HGNC Gene Table (the Anchor)

The gene-centric anchor is [HGNC](https://www.genenames.org/)'s complete
gene set - the single authoritative source for the current approved
symbol *and* every previous/alias symbol a gene has ever had, alongside
its Ensembl gene ID, UniProt accession(s), and Entrez ID, all in one row
per approved gene (~19,200 of them are protein-coding).
`getHgncGeneTable()` downloads (and caches) it and reshapes the
pipe-separated multi-value columns into list-columns:

```{r hgnc_table, eval=FALSE}
hgncTable <- getHgncGeneTable()
nrow(hgncTable)
hgncTable[hgncTable$symbol == "FGFR1", c("symbol", "ensembl_gene_id")]
hgncTable$uniprot_ids[hgncTable$symbol == "FGFR1"]
```

By default this uses a **pinned quarterly HGNC snapshot** (not the
rolling always-latest file) specifically so a master table built today
and one built next year start from the same gene list unless you
deliberately ask otherwise (`current = TRUE` for the rolling file, or
`archiveFile = "..."` for any other specific HGNC archive snapshot).

**Symbol normalization matters here**: DGIdb and TTD echo back whatever
gene symbol was current when *their own* records were curated, and a
nontrivial fraction are now outdated relative to HGNC. `buildHgncSymbolMap()`
builds a lookup from every historical `prev_symbol`/`alias_symbol` to its
gene's current approved symbol; `normalizeGeneSymbols()` applies it
(flagging any symbol it still can't resolve, rather than silently
dropping it):

```{r hgnc_symbol_map, eval=FALSE}
symbolMap <- buildHgncSymbolMap(hgncTable)
normalizeGeneSymbols(c("FGFR1", "ABL"), symbolMap = symbolMap, hgncTable = hgncTable)
```

## Building the Table, With Checkpointing

`buildGenomeWideDrugTargetTable()` loops the four annotation sources'
own bidirectional functions over every gene in an HGNC table, in
checkpointed chunks written to `outDir` as it goes - so an interrupted
run resumes from the last completed chunk instead of starting over
(`rerun = FALSE`, the default, skips anything the chunk manifest already
marks done):

```{r genomewide_build, eval=FALSE}
ttdDbPath <- buildTtdDb(rerun = FALSE)
masterTable <- buildGenomeWideDrugTargetTable(
    hgncTable = hgncTable,
    sources = c("chembl", "dgidb", "opentargets", "ttd"),
    ttdDbPath = ttdDbPath,
    outDir = "genomewide_build",  # persists chunk files + a manifest here
    chunkGenes = 500L)
sapply(masterTable, nrow)
```

This deliberately covers only the four **curated-annotation** sources -
not ChEMBL's bioassay function or PubChem (see [Bioassay Queries]).
ChEMBL's full activity table is tens of millions of rows and PubChem's
accessor is inherently a one-gene-at-a-time loop; a genome-wide sweep of
either is a different kind of job, better done from their bulk data
downloads than by looping this package's live-API functions tens of
thousands of times.

The result is a named list - one data.frame per source, each row tagged
with `hgnc_id`/`symbol`/`ensembl_gene_id` alongside that source's own
native columns (deliberately *not* passed through `combineDrugTargets()`'s
harmonization, so no information is lost from a resource meant to be
reused repeatedly; apply `combineDrugTargets()` afterward if you want the
flattened cross-source view). The assembled result is also cached via
`BiocFileCache`, so a completed build is reusable across sessions and
machines without recomputation - exactly the "compute every few months,
share the result" workflow this section is meant to support.

# Supplement

The functions below predate, and are independent of, the sources and
translation layer described above. They remain available as
backward-compatible/alternative options but are not the default workflow
recommended by this vignette.

## Local ChEMBL SQLite Workflow

Before ChEMBL's REST API was added, this package's ChEMBL support was
exclusively through a downloaded local SQLite instance of the full ChEMBL
database, queried via `drugTargetAnnot()` (and the related
`getDrugTarget()`, which queries a pre-generated flat-file export rather
than the SQLite directly). This remains useful for bulk, offline, or
very-high-volume ChEMBL querying where the REST API's batch limits (see
[Performance Considerations]) are impractical.

```{r legacy_chembl_download, eval=FALSE}
config <- genConfig(chemblDbPath = "chembldb.db")
downloadChemblDb(rerun = TRUE, config = config)  # downloads the full ChEMBL SQLite (several GB)
```

The examples below instead point `chemblDbPath` at the small ChEMBL
subset shipped with the package (`inst/extdata/chembl_sample.db`), so
they run - and are checked - as part of every vignette build rather than
only being documented. `drugTargetAnnot()` additionally needs a compound
ID cross-reference table built once via `downloadUniChem()`/
`cmpIdMapping()` (see [Static UniChem Mirror] below) - `resultsPath` is
pointed at a temp directory here purely to avoid leaving files behind in
the vignette source tree:

```{r legacy_chembl_sample, eval=TRUE}
config <- genConfig(
    chemblDbPath = system.file("extdata", "chembl_sample.db", package = "drugTargetInteractions"),
    resultsPath = tempfile())
downloadUniChem(config = config, rerun = TRUE)
cmpIdMapping(config = config, rerun = TRUE)
t2d <- drugTargetAnnot(
    list(molType = "protein", idType = "UniProt_ID", ids = "P00915"), config = config)
head(t2d[, c("Drug_Name", "MOA", "Action_Type", "First_Approval")])
d2t <- drugTargetAnnot(
    list(molType = "cmp", idType = "chembl_id", ids = "CHEMBL25"), config = config)  # aspirin
d2t[, c("UniProt_ID", "Organism", "MOA")]
```

`getDrugTarget()` queries a pre-generated flat-file export instead of the
SQLite directly - built once with `drugTargetAnnotTable()` and re-read on
every subsequent call, which is faster for repeated querying against the
same ChEMBL snapshot:

```{r legacy_getdrugtarget, eval=TRUE}
annotFile <- tempfile(fileext = ".xls")
drugTargetAnnotTable(outfile = annotFile, config = config, rerun = TRUE)
id_mapping <- c(chembl = "chembl_id", pubchem = "PubChem_ID",
    uniprot = "UniProt_ID", drugbank = "DrugBank_ID")
getDrugTarget(dt_file = annotFile,
    queryBy = list(molType = "cmp", idType = "chembl", ids = "CHEMBL25"),
    id_mapping = id_mapping)[, c("pref_name", "action_type", "UniProt_ID", "Organism")]
```

`drugTargetBioactivity()` returns the corresponding raw bioassay
measurements (IC50/Ki/Kd/EC50) from the same local SQLite:

```{r legacy_bioactivity, eval=TRUE}
ba <- drugTargetBioactivity(
    list(molType = "protein", idType = "uniprot", ids = "P00915"), config = config)
head(ba[, c("pref_name", "standard_type", "standard_value", "standard_units")])
```

## UniProt.ws-Based ID Mapping

`getUniprotIDs()` predates `getUniprotMapping()` (see [ID Translation
Layer]) and is based on the Bioconductor `r Biocpkg("UniProt.ws")`
package rather than a direct REST call. It is kept for backward
compatibility; `getUniprotMapping()` is the recommended path going
forward.

```{r legacy_uniprot_ws, eval=FALSE}
getUniprotIDs(taxId = 9606, kt = "ENSEMBL", keys = c("ENSG00000077782"))
```

## biomaRt-Based Paralogs

`getParalogs()` predates `getEnsemblParalogs()`/`getEnsemblOrthologs()`
(see [ID Translation Layer]) and is based on `r CRANpkg("biomaRt")`'s
BioMart interface rather than Ensembl's REST API directly. It returns
within-human paralogs only (no cross-species orthologs). Kept for
backward compatibility.

```{r legacy_paralogs, eval=FALSE}
queryBy <- list(molType = "gene", idType = "external_gene_name", ids = "FGFR1")
getParalogs(queryBy)
```

## ensembldb-Based ID Mapping

`getSymEnsUp()` is a third, independent way to translate between gene
symbols, Ensembl gene IDs and UniProt accessions - based on a local
Bioconductor annotation package (`r Biocpkg("EnsDb.Hsapiens.v86")`) via
`r Biocpkg("ensembldb")`, rather than a live web service. Useful when
network access to UniProt's or Ensembl's REST endpoints isn't available.
Supported `idtype` values are `GENE_NAME`, `ENSEMBL_GENE_ID` and
`UNIPROT_ID`.

```{r legacy_symensup, eval=TRUE}
idMap <- getSymEnsUp(EnsDb = "EnsDb.Hsapiens.v86", ids = "FGFR1", idtype = "GENE_NAME")
head(idMap$idDF)
```

## Static UniChem Mirror

`downloadUniChem()`/`cmpIdMapping()` predate `buildUnichemDb()`/
`getUnichemMapping()` (see [ID Translation Layer]) and work from a small,
static, pre-computed ChEMBL&harr;{DrugBank, PubChem, ChEBI} cross-reference
snapshot hosted on a project S3 bucket, rather than UniChem's own current
bulk data. Kept for backward compatibility and for the local-ChEMBL-SQLite
workflow above, which uses it internally via `cmpIdMapping()`.

```{r legacy_unichem, eval=FALSE}
downloadUniChem(rerun = TRUE)
cmpIdMapping(outfile = "cmp_ids.rds", rerun = TRUE)
```

## Putting It Together

`runDrugTarget_Annot_Bioassay()` is a convenience wrapper combining
`drugTargetAnnot()` and `drugTargetBioactivity()` for a set of UniProt
IDs obtained from `getUniprotIDs()` or `getParalogs()` - the legacy
equivalent of this vignette's own [Cross-Source Queries] combining step,
predating `combineDrugTargets()`. It expects the two-slot (`IDM`/`SSNN`)
list structure returned by those ID-mapping functions:

```{r legacy_run_everything, eval=TRUE}
idMap <- getSymEnsUp(EnsDb = "EnsDb.Hsapiens.v86", ids = c("CA7", "CFTR"), idtype = "GENE_NAME")
queryBy <- list(molType = "gene", idType = "ensembl_gene_id", ids = names(idMap$ens_gene_id))
res_list <- getParalogs(queryBy)
combined <- runDrugTarget_Annot_Bioassay(res_list = res_list, up_col_id = "ID_up_sp",
    ens_gene_id = idMap$ens_gene_id, config = config)
sapply(combined, nrow)
head(combined$Annotation[, c("GeneName", "Drug_Name", "MOA")])
```

# Session Info

```{r sessionInfo}
sessionInfo()
```

# References
